---
title: "Preliminary work"
output:
  html_document:
    keep_md: false
    toc: true
    toc_float: true
    toc_depth: 4

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup

```{r libraries}
## library(FirebrowseR)
## library(BiocInstaller)
library(biomaRt)
library(knitr)
## library(rlist)
library(edgeR)
## library(statmod)
library(org.Hs.eg.db)
library(RSQLite) #
library(IRanges)
```

```{r settings}
opts_chunk$set(fig.width = 12,
               fig.height = 12,
               cache = TRUE,
               include = TRUE,
               cache.lazy = FALSE,
               warning = TRUE,
               message = TRUE)
```

# TFBS data retrieval
For transfac + fimo scans
### Fimo download:
```{bash, eval = TRUE}
mkdir -p utils 

cd $_
# retrieving transfac+fimo
wget -q https://noble.gs.washington.edu/custom-tracks/fimo.hg19.transfac-0.1.bb

# getting the exec for linux 64 bits

rsync -aP rsync://hgdownload.soe.ucsc.edu/genome/admin/exe/linux.x86_64/bigBedToBed .

# making it runnable

chmod a+x bigBedToBed

# converting to bed

./bigBedToBed fimo.hg19.transfac-0.1.bb test.bed

# checking how a bedfile looks like

head test.bed

cut -f4 test.bed | sort | uniq -c | sed 's/^\s*//' > fimo_transfac.txt
mv fimo_transfac.txt ../data

```
### Adding the Fimo list into R
```{r}
fimo_transfac <- unique(read.table(file.path('data', 'fimo_transfac.txt'),
                                header = FALSE,
                                stringsAsFactors = FALSE))
colnames(fimo_transfac)<- c("Number", "tf")
```

# RSEM download:
                          CAN THIS CHUNKS BE DONE IN A FUNCTION??
      The idea I have is to leave the BRCA like these and then do the rest on a function maybe?

### BRCA
```{r brca_download}
download.file('http://gdac.broadinstitute.org/runs/stddata__2016_01_28/data/BRCA/20160128/gdac.broadinstitute.org_BRCA.Merge_rnaseqv2__illuminahiseq_rnaseqv2__unc_edu__Level_3__RSEM_genes_normalized__data.Level_3.2016012800.0.0.tar.gz',  destfile = file.path(WD, 'brca_rsem.tar.gz'))

brca.fn <- grep('data.txt', untar(file.path(WD, 'brca_rsem.tar.gz'), list = TRUE), value = TRUE)
untar(file.path(WD, 'brca_rsem.tar.gz'), files = brca.fn)

header.brca <- read.table(file = brca.fn, sep = '\t', header = FALSE, nrow = 1,
                     stringsAsFactors = FALSE)

brca <- read.table(file = brca.fn, sep = '\t', header = TRUE, skip = 2)
colnames(brca) <- header.brca[1,]

rownames(brca) <- brca[,1]
brca <- brca[,-1]
rm(header.brca)
```

### COAD
```{r coad_download}
download.file('http://gdac.broadinstitute.org/runs/stddata__2016_01_28/data/COAD/20160128/gdac.broadinstitute.org_COAD.Merge_rnaseqv2__illuminahiseq_rnaseqv2__unc_edu__Level_3__RSEM_genes_normalized__data.Level_3.2016012800.0.0.tar.gz',  destfile = file.path(WD, 'coad_rsem.tar.gz'))

coad.fn <- grep('data.txt', untar(file.path(WD, 'coad_rsem.tar.gz'), list = TRUE), value = TRUE)
untar(file.path(WD, 'coad_rsem.tar.gz'), files = coad.fn)

header.coad <- read.table(file = coad.fn, sep = '\t', header = FALSE, nrow = 1,
                     stringsAsFactors = FALSE)

coad <- read.table(file = coad.fn, sep = '\t', header = TRUE, skip = 2)
colnames(coad) <- header.coad[1,]

rownames(coad) <- coad[,1]
coad <- coad[,-1]
rm(header.coad)
```

### LUAD:
```{r luad_download}
download.file('http://gdac.broadinstitute.org/runs/stddata__2016_01_28/data/LUAD/20160128/gdac.broadinstitute.org_LUAD.Merge_rnaseqv2__illuminahiseq_rnaseqv2__unc_edu__Level_3__RSEM_genes_normalized__data.Level_3.2016012800.0.0.tar.gz',  destfile = file.path(WD, 'luad_rsem.tar.gz'))

luad.fn <- grep('data.txt', untar(file.path(WD, 'luad_rsem.tar.gz'), list = TRUE), value = TRUE)
untar(file.path(WD, 'luad_rsem.tar.gz'), files = luad.fn)

header.luad <- read.table(file = luad.fn, sep = '\t', header = FALSE, nrow = 1,
                     stringsAsFactors = FALSE)

luad <- read.table(file = luad.fn, sep = '\t', header = TRUE, skip = 2)
colnames(luad) <- header.luad[1,]

rownames(luad) <- luad[,1]
luad <- luad[,-1]
rm(header.luad)
```

### PRAD
```{r prad_download}
download.file('http://gdac.broadinstitute.org/runs/stddata__2016_01_28/data/PRAD/20160128/gdac.broadinstitute.org_PRAD.Merge_rnaseqv2__illuminahiseq_rnaseqv2__unc_edu__Level_3__RSEM_genes_normalized__data.Level_3.2016012800.0.0.tar.gz',  destfile = file.path(WD, 'prad_rsem.tar.gz'))

prad.fn <- grep('data.txt', untar(file.path(WD, 'prad_rsem.tar.gz'), list = TRUE), value = TRUE)
untar(file.path(WD, 'prad_rsem.tar.gz'), files = prad.fn)

header.prad <- read.table(file = prad.fn, sep = '\t', header = FALSE, nrow = 1,
                     stringsAsFactors = FALSE)

prad <- read.table(file = prad.fn, sep = '\t', header = TRUE, skip = 2)
colnames(prad) <- header.prad[1,]

rownames(prad) <- prad[,1]
prad <- prad[,-1]
rm(header.prad)
```
# Mapping

### Symbols

We would like to transform TRANSFAC names to HUGO symbols (we'll check differential expression of these using TCGA RSEM data)
Transformation of Fimo list (TRANSFAC names) to match our TCGA RSEM data:
```{r mapping}
con <- org.Hs.eg_dbconn()

stmt <- 'SELECT * FROM alias, gene_info WHERE alias._id == gene_info._id;'
# execute the query on the database
aliases <- dbGetQuery(con, stmt)

candidates <- aliases[aliases$alias_symbol %in% toupper(fimo_transfac$tf),]
candidates2 <- aliases[aliases$symbol %in% toupper(fimo_transfac$tf),]

valid_symbols <- unique(candidates$symbol, candidates2$symbol)
```

### Entrez IDs
Gene symbols to Entrez gene ids (TCGA RSEMs also provide these identifiers)

```{r biomart_entrez}
mart <- useMart("ENSEMBL_MART_ENSEMBL", host="grch37.ensembl.org",
                path="/biomart/martservice", dataset="hsapiens_gene_ensembl")


entrez_tf <- getBM(attributes=c('hgnc_symbol', 
                                'entrezgene'), 
                   filters = 'external_gene_name', 
                   values = valid_symbols,
                   mart = mart)

```

### Defining the mapping function:

Getting a dictionary of TF with their TCGA id, Entrez id and standard gene symbol for which we know the TRANSFAC/FIMO motif.
          TO DO WITH A GENERIC FUNCTION, SO WE CAN USE IN THE IN A LITTLE CHUNK FOR EACH TCGA DATA.
          
```{r}
mapping <- data.frame(tcga = rownames(coad),
                      entrez = sapply(strsplit(rownames(coad), '|', fixed = TRUE),
                                      function(x) return(x[2])),
                      symbol = sapply(strsplit(rownames(coad), '|', fixed = TRUE),
                                      function(x) return(x[1])))

tf <- merge(mapping, entrez_tf, by.x = 'entrez', by.y = 'entrezgene')


## so tf$tcga will contain the TCGA identifiers (rownames of COAD)
(coad[tf$tcga,])

```


# Analysis:

I have in mind we can try to compare between normal and tumor and then compare between participants, lets see the preliminary results what they show:

## Diff expression:

###Normal vs tumor, paired mode:


Getting sort of counts from RSEMs (this is not correct because we are just rounding RSEMs!) and generating a DGE object only for those samples that come from the same patient, because we will compare tumors against matched normals.

```{r}
coad.r <- round(coad) ## this is dirty, they are RSEMs and not counts!!!!!!!

coad.data <- data.frame(barcode = colnames(coad),
                      sample = strtrim(sapply(strsplit(colnames(coad), '-', fixed = TRUE),
                                              function(x) return(x[4])), 2),
                      participant = sapply(strsplit(colnames(coad), '-', fixed = TRUE),
                                              function(x) return(x[3]))
                      )
#Participants that have normal vs tumor tissue comparision:
head(coad.data)
dim(coad.data)
table(coad.data$sample)
## only normals and primary tumors
coad.data.1 <- coad.data[coad.data$sample %in% c('01', '11'),]

## only matched normal and tumor from the same participant
selected <- coad.data.1$participant[duplicated(coad.data.1$participant)]
coad.data.2 <- coad.data.1[coad.data.1$participant %in% selected,]
coad.barcode <- coad[,coad.data.2$barcode]

coad_dge <- DGEList(coad.barcode, samples = coad.data.2)
```

Norm factors
```{r}
coad_dge_norm <- calcNormFactors(coad_dge)
```

```{r}
plotMDS(coad_dge_norm)
```

Desing the matrix for the analysis:
```{r}
Normal<-factor(c("11"== coad_dge_norm$samples[5]))
#Tumor<-factor(c("01"== coad_dge_norm$samples[5]))

matrix.model<-model.matrix(~Normal)

rownames(matrix.model)<-colnames(coad_dge_norm$counts)
head(matrix.model)
dim(matrix.model)
```
Do tumors match tumors and normals match normals? Yes they do, confirmed but only in the general mode.

Or do samples from the same participant cluster together? I thought of the possibility but i dont know if that gives us additional information, we dont have other information about the participant we could use to, like if he/she is dead or alive or other kind of info, like weight, smoker or not (LUAD), etc...

What does this mean (biologically)? 

```{r}
xy.coad.disp<-estimateDisp(coad_dge_norm,matrix.model,robust = FALSE)
#Says it is redundant the tumor column so we erase it
xy.coad.disp$common.dispersion
```

```{r}
plotBCV(xy.coad.disp)
```

Differential expression:
```{r}
coad.fit<-glmFit(xy.coad.disp, matrix.model)
coad.lrt<-glmLRT(coad.fit)
topTags(coad.lrt)
```

```{r}
o.coad <- order(coad.lrt$table$PValue)
coad.cpm<-cpm(xy.coad.disp)[o.coad[1:10],]
head(colnames(coad.cpm))
```

```{r}
summary(decideTests(coad.lrt))
```

```{r}
plotMD(coad.lrt)
abline(h=c(-1, 1), col="green")
```